Study sample characteristics in chronobiology and sleep research

### 2021-06-14 article.Rmd
Update article.Rmd
Edit typos.

authornote: |
 This work was funded by the Wellcome Trust (Sir Henry Wellcome Fellowship to MS 204686/Z/16/Z; Research Enrichment – Diversity & Inclusion WT 204686/Z/16/A) and Linacre College (Biomedical Sciences Junior Research Fellowship to MS).
abstract: |
  Many aspects of sleep and circadian physiology appear to be sensitive to participant-level characteristics. While recent research robustly highlights the importance of considering participant-level demographic information, it is not clear to what extent this information is available in the large body of already published literature. Here, we investigated study sample characteristics in the published sleep and chronobiology research over the past 40 years. 6,777 articles were identified and a random sample of 20% was included. The reporting of sample size, age, sex, gender, ethnicity, level of education, socio-economic status, and profession of the study population was scored, and any reported aggregate summary statistics for these variables were recorded. We found that while >90% of studies reported age or sex, all other variables were reported in <10% of cases. Sex balance greatly changed over the years, from a ~3:1 male to female ratio in the 1990s to a near-equal representation in the 2010s. We found that the majority of studies report at least sex or age, while other variables are typically not reported. Reporting quality is highly variable, indicating an opportunity to standardize reporting guidelines for participant-level characteristics to facilitate meta analyses.

@@

In this article, we addressed the question how representiative, diverse and inclusive published research articles in sleep and chronobiology are. We analysed a sample of >1,300 articles published in sleep research and chronobiology journals between 1979 and 2019 for its inclusion, reporting analysis of demographic variables of study participants, including age, sex, gender, ethnicity, level of education, socio-economic status, and professio.  We found that while >90% of studies reported age or sex, all other variables were reported in <10% of cases, with the frequency of reporting changing over time. We identify opportunities for improving the reporting of demographic variables.

@@

Future research needs to identify:
1. Schemas for reporting demographic variables in a harmonised way across geographical and culutral contexts;
2. Gaps in the sleep and chronobiology literature with respect to understudied populations;
3. Understand the extent to which research practices run counter the inclusion of diverse populations in research in all stages of the research cycle, and how this can exacerbate health inequities.

@@

There are large individual differences in and circadian physiology [@van_dongen_individual_2005; @kerkhof_inter-individual_1985; @tankova_circadian_1994; @horne_individual_1977; @baehr_individual_2000; @burgess_individual_2008; @santhi_spectral_2012; @chellappa_individual_2021], demonstrating the need to consider sleep and circadian physiology at participant-level. Some of these characteristics are systematically related to demographic variables, most notably sex [@redline_effects_2004; @santhi_sex_2016; @cain_sex_2010; @anderson_sexual_2020; @mong_sleep_2011], age [@redline_effects_2004; @espiritu_aging-related_2008; @mander_sleep_2017; @desforges_sleep_1990; @li_sleep_2018; @bliwise_sleep_1993; @benloucif_responsiveness_2006; @duffy_aging_2015] and ethnicity [@ahn_scoping_2021; @goldstein_sleep_2020; @eastman_blacks_2012; @eastman_circadian_2016]. Zooming out from the level of individual studies, there are significant inequities in sleep health depending on ethnicity and socio-economic status [@jackson_sleep_2020; @jackson_racialethnic_2020; @williams_racialethnic_2015; @jackson_workshop_2020; @jean-louis_importance_2016; @laposky_reducing_2016]. As compromised sleep has many knock-on effects, including negative effects on cardiovascular, metabolic, neurobehavioural and cognitive function, it is a key imperative to understand how demographic variables influence sleep and circadian rhythms.

The extent to which a scientific field's findings are generalisable are a function of the representativeness of a given study sample, and the extent to which findings differ between, e.g., different sexes and ethnicity In the domain of sex, a recent study reviewed the reporting and analysis of sex in biological sciences research [@woitowich_10-year_2020]. The authors found that while sex inclusion has significantly increased over the past 10 years [@beery_sex_2011], sex-based analysis has not improved, despite recent policies and funder mandates [@clayton_policy_2014]. The term "gender data gap" has recently been introduced, demonstrating that women have historically been excluded from biomedical research [@criado-perez_invisible_2020]. Similarly, the lack of diversity in biomedical and clinical research, and understudy of minorities in the presence of existing health inequities exacerbates these inequities [@oh_diversity_2015].

@@

6,777 articles were identified through a MEDLINE search by the journal and including odd years. A random sample of 20% was initially selected for screening. Inclusion requirements included conducting original research in the English language, reporting human data, and recruiting volunteers. As such, animal studies, bibliographies, case reports, comments, conference proceedings, editorials, guidelines, letters, retracted publications, reviews, errata and corrigenda were excluded.

@@

## Number of analysed articles

Out of `r as.numeric(nrow(data.original))` identified and pre-screened articles, we included and extracted data from `r as.numeric(nrow(data.included))` (`r round(percent.included)`%). The distribution of years in which the articles were published is non-uniform and we included and extracted data from more recent articles (Fig. 1). In addition, the representation of journals in the final list of articles is non-uniform, not least as the included journals will have not have been available from the entire data collection period (1979 and onwards).

```{r, fig.cap="Included and analysed articles by year and journal. More recent articles are more represented, reflecting an overall increase in scientific output.", warning=FALSE, message=FALSE}
figure1
```

We also examined reasons for exclusion amongst the articles that we did not include and extract data from. These are given in Figure 2. Reasons for exclusion vary somewhat between different years, with articles not reporting original research being excluded at the highest rate on average.

@@

  We examined whether sample size reported in studies, and the sample sizes reported. Overall, sample size was reported in `r round(percent.ss)`% of studies. We furthermore examined the distribution of samples sizes as a function of the publication year of the article (Fig. 5), showing a wider distribution of sample sizes in more recent articles.

@@

  We examined the reporting of the demographic variables ethnicity, education, profession and socio-economic status. These demographics variables were reported in `r round(percent.education)`% of studies for education, `r round(percent.ethnicity)`% for ethnicity, `r round(percent.profession)`% for profession, and `r round(percent.ses)`% for socio-economic status. Figure 8 shows the distribution of this reporting across the years. Qualitatively, there is a clear increase of reporting additional demographic variables over time, with ethnicity being the most reported demographic variables across articles in more recent years.

```{r, fig.cap="test", warning=FALSE, message=FALSE}
figure10
```
  We furthermore examined the number of categories included for each of these demographic variables. In Figure 9, we show the number of categories reported for each variable among those articles that included it in a histogram. On overage, we find that the coding scheme for each of these variables in the included articles comprised between 3 and 4 variables.

@@

  We examined the extent to which articles reported subgroup analyses of the data based on one or more of the reported demographic variables. Over time (Fig. 9), we see a distinct evolution of the extent to which subgroup analyses of the study sample were performed. The most common subgroup analyses involve disaggregting by sex and age, or both.

@@

Here, we considered the inclusion, reporting of `r as.numeric(nrow(data.included))` articles in chronobiology and sleep research, sampled from eight journals. More articles from more recent years are included, with articles originating in the US representing the large group. Over time, we found an increase in the range of sample sizes. We furthermore find a wider spread of average ages included in articles. Across the included articles, we observe an increase in reporting of demographic variables (ethnicity, education, professional and socioeconomic status) over time, with ethnicity being the most-reported variable in recent years. Of studies included these demographic variables, the number of categories varies widely between studies.

@@

The ability to generalise findings from the scientific literature to wide and diverse populations of people hinges upon the representativeness of the study sample with respect to demographic categories. The question to what extent the composition of a given study sample can make the generalisability of findings difficult or impossible has received attention in the field of psychology, where many articles published in prominent journals reflected participants from WEIRD (Western, Educated, Industrialized, Rich, and Democratic) contexts [@henrich_weirdest_2010; @muthukrishna_beyond_2020]. In other fields, analyses similar to the one in the present review have been published [@obryant_reporting_2004; @sifers_reporting_2002; @jones_reporting_2020], but to our knowledge, this review represents a first look at the inclusion, reporting and analysis participant demographics in chronobiology and sleep research.

@@

There is convergent knowledge that sleep and circadian physiology and health are subject to large individual differences. Demographic variables provide on lens through which to understand individual differences, and importantly can provide insights into systemic disadvantages and inequities. In the clinical domain, the need to time therapy based on a patient's individual circadian rhythm has more recently become the focus of the emerging field of chronotherapy or chronotherapeutics [@adam_core_2019; @greco_personalized_2020; @dijk_novel_2020; @hill_optimizing_2020]. Understanding interindividual variability needs to become a key research area to understand the circadian and sleep physiology in the face of human diversity.

@@

We turn to possible limitations of this review and the included analyses and discuss how they might introduce bias in our findings. First, we consider the possibility that the article selection procedure may have introduced biases. Our review only concerned articles from a subset of eight specialized journals. As a consequence, the included articles were necessarily published in these journals, ignoring relevant articles published in other specialised journals (such as those included in the list of candidate journals), and articles published in other, including interdisciplinary journals. This raises the question to what extent we may have missed a section of the literature that would have been relevant to include here. As an alternative strategy, we considered randomly sampling a subset of chronobiology and sleep research articles produced by a general search (e.g. on search from "sleep OR chronobiology" on MEDLINE), but considered this to be too permissive. Our strategy of selecting a subset of candidate journals provided a reasonable trade-off, as well as sampling from a range of field-specific outlets.

Due to the non-uniform distribution of publication years of the included articles (Fig. 1), variables derived from published papers and visualized and/or analyzed by year will have varying uncertainty, with reported percentages from publications of the earlier years being most uncertain. The fact that early years are represented with fewer articles, however, is a not a function of our data set, but of the exponential growth of scientific output [@powell_introduction_2017; @bornmann_growth_2015; @parolo_attention_2015].

## Towards standardised reporting of demographic variables: From checklists to schemas?

There are guidelines and/or checklists for standardizing reporting of participant characteristics, such as CONSORT [@schulz_consort_2010] or STROBE [@von_elm_strengthening_2007] (an extensive data base for health research reporting guidelines is provided by the Equator Network, https://www.equator-network.org/). Some biomedical journals [e.g. @robinson_transparent_2017] specifically state demographic reporting requirements in the author instructions. Similarly, some organisations may make recommendations of specific reporting items for specific types of study questions [@veitch_cie_2020].

These guidelines and/or checklists are largely focused on *what* is reported and not *how* it is reported. There is, *a priori*, however, no reason to not develop and use a standardized and machine-readable schema for reporting participant characteristics. The FAIR principles state that data should be findable, accessible, interoperable and reusable [@wilkinson_fair_2016], and one way of realising these criteria is the use of data schemas which could prescribe categories of data and common naming schemes for reporting participant characteristics. Importantly, however, "what gets counted counts" [@dignazio_data_2020], and it will be imperative to understand to what extend such any data schema may be exclusionary, e.g., by enforcing gender binaries [@hyde_future_2019], and whether any specific demographic variable is truly important (following the principle of data minimization). A further central point to consider is the extent to which disaggregation by demographic variables could be used to do harm in some way [@dignazio_data_2020].

@@

This review provided a first look at the inclusion, reporting and analysis of demographic variables in the chronobiology and sleep research literature, considering >1000 articles across eight journals. We addressed the need to consider individual differences, as well as the dependence of sleep and circadian rhythms ond demographic variables. Additionally, we we outline an opportunity to improve reporting of participant-level characteristics using formalised data schemas.

>>>>

authornote: |
 This work was funded by the Wellcome Trust (Sir Henry Wellcome Fellowship to MS 204686/Z/16/Z; Research Enrichment – Diversity & Inclusion WT 204686/Z/16/A) and Linacre College (Biomedical Sciences Junior Research Fellowship to MS).
abstract: |
  Many aspects of sleep and circadian physiology appear to be sensitive to participant-level characteristics. While recent research robustly highlights the importance of considering participant-level demographic information, it is not clear to what extent this information is available in the large body of already published literature. Here, we investigated study sample characteristics in the published sleep and chronobiology research over the past 40 years. 6,777 articles were identified and a random sample of 20% was included. The reporting of sample size, age, sex, gender, ethnicity, level of education, socio-economic status, and profession of the study population was scored, and any reported aggregate summary statistics for these variables were recorded. We found that while >90% of studies reported age or sex, all other variables were reported in <15% of cases. Sex balance greatly changed over the years, from a ~3:1 male to female ratio in the 1990s to a near-equal representation in the 2010s. We found that the majority of studies report at least sex or age, while other variables are typically not reported. Reporting quality is highly variable, indicating an opportunity to standardize reporting guidelines for participant-level characteristics to facilitate meta analyses.

@@

In this article, we addressed the question of how representative, diverse and inclusive published research articles in sleep and chronobiology are. We analyzed a sample of >1,300 articles published in sleep research and chronobiology journals between 1979 and 2019 for its inclusion, reporting analysis of demographic variables of study participants, including age, sex, gender, ethnicity, level of education, socio-economic status, and profession.  We found that while >90% of studies reported age or sex, all other variables were reported in <10% of cases, with the frequency of reporting changing over time. We identify opportunities for improving the reporting of demographic variables.

@@

Future research needs to identify:
1. Schemas for reporting demographic variables in a harmonized way across geographical and cultural contexts;
2. Gaps in the sleep and chronobiology literature with respect to understudied populations;
3. Understand the extent to which research practices run counter the inclusion of diverse populations in research in all stages of the research cycle, and how this can exacerbate health inequities.

@@

There are large individual differences in sleep and circadian physiology [@van_dongen_individual_2005; @kerkhof_inter-individual_1985; @tankova_circadian_1994; @horne_individual_1977; @baehr_individual_2000; @burgess_individual_2008; @santhi_spectral_2012; @chellappa_individual_2021], demonstrating the need to consider sleep and circadian physiology at a participant-level. Some of these characteristics are systematically related to demographic variables, most notably sex [@redline_effects_2004; @santhi_sex_2016; @cain_sex_2010; @anderson_sexual_2020; @mong_sleep_2011], age [@redline_effects_2004; @espiritu_aging-related_2008; @mander_sleep_2017; @desforges_sleep_1990; @li_sleep_2018; @bliwise_sleep_1993; @benloucif_responsiveness_2006; @duffy_aging_2015] and ethnicity [@ahn_scoping_2021; @goldstein_sleep_2020; @eastman_blacks_2012; @eastman_circadian_2016]. Zooming out from the level of individual studies, there are significant inequities in sleep health depending on ethnicity and socio-economic status [@jackson_sleep_2020; @jackson_racialethnic_2020; @williams_racialethnic_2015; @jackson_workshop_2020; @jean-louis_importance_2016; @laposky_reducing_2016]. As compromised sleep has many knock-on effects, including negative effects on cardiovascular, metabolic, neurobehavioral and cognitive function, it is a key imperative to understand how demographic variables influence sleep and circadian rhythms.

The extent to which a scientific field's findings are generalizable is a function of the representativeness of a given study sample, and the degree to which findings differ between, e.g., different sexes and ethnicity. In the domain of sex, a recent study reviewed the reporting and analysis of sex in biological sciences research [@woitowich_10-year_2020]. The authors found that while sex inclusion has significantly increased over the past 10 years [@beery_sex_2011], sex-based analysis has not improved, despite recent policies and funder mandates [@clayton_policy_2014]. The term "gender data gap" has recently been introduced, demonstrating that women have historically been excluded from biomedical research [@criado-perez_invisible_2020]. Similarly, the lack of diversity in biomedical and clinical research, and understudy of minorities in the presence of existing health inequities exacerbates these inequities [@oh_diversity_2015].

@@

6,777 articles were identified through a MEDLINE search and filtering by journal and odd years. A random sample of 20% was initially selected for screening. Inclusion requirements included conducting original research in the English language, reporting human data, and recruiting volunteers. As such, animal studies, bibliographies, case reports, comments, conference proceedings, editorials, guidelines, letters, retracted publications, reviews, errata and corrigenda were excluded.

@@

## Number of analyzed articles

Out of `r as.numeric(nrow(data.original))` identified and pre-screened articles, we included and extracted data from `r as.numeric(nrow(data.included))` (`r round(percent.included)`%). The distribution of years in which the articles were published is non-uniform and we included and extracted data from more recent articles (Fig. 1). In addition, the representation of journals in the final list of articles is non-uniform, not least as the included journals will not have been available for the entire data collection period (1979 and onwards).

```{r, fig.cap="Included and analyzed articles by year and journal. More recent articles are more represented, reflecting an overall increase in scientific output.", warning=FALSE, message=FALSE}
figure1
```

We also examined reasons for exclusion among the articles that we did not include and extract data from. These are given in Figure 2. Reasons for exclusion vary somewhat between different years, with articles not reporting original research being excluded at the highest rate on average.

@@

  We examined whether sample size was reported in studies, and the sample sizes reported. Overall, sample size was reported in `r round(percent.ss)`% of studies. We furthermore examined the distribution of samples sizes as a function of the publication year of the article (Fig. 5), showing a wider distribution of sample sizes in more recent articles.

@@

  We examined the reporting of the demographic variables ethnicity, education, profession and socio-economic status. These demographics variables were reported in `r round(percent.education)`% of studies for education, `r round(percent.ethnicity)`% for ethnicity, `r round(percent.profession)`% for profession, and `r round(percent.ses)`% for socio-economic status. Figure 8 shows the distribution of this reporting across the years. Qualitatively, there is a clear increase in the reporting of additional demographic variables over time, with ethnicity being the most reported demographic variable across articles in more recent years.

```{r, fig.cap="test", warning=FALSE, message=FALSE}
figure10
```
  We furthermore examined the number of categories included for each of these demographic variables. In Figure 9, we show the number of categories reported for each variable among those articles that included it in a histogram. On overage, we find that the coding scheme for each of these variables in the included articles comprised between three and four variables.

@@

  We examined the extent to which articles reported subgroup analyses of the data based on one or more of the reported demographic variables. Over time (Fig. 9), we see a distinct evolution of the extent to which subgroup analyses of the study sample were performed. The most common subgroup analyses involve disaggregating by sex, age, or both.

@@

Here, we considered the inclusion, reporting of `r as.numeric(nrow(data.included))` articles in chronobiology and sleep research, sampled from eight journals. More articles from more recent years are included, with articles originating in the US representing the large group. Over time, we found an increase in the range of sample sizes. We furthermore found a wider spread of average ages included in articles. Across the included articles, we observed an increase in the reporting of demographic variables (ethnicity, education, professional and socioeconomic status) over time, with ethnicity being the most-reported variable in recent years. Of studies included these demographic variables, the number of categories varied widely between studies.

@@

The ability to generalize findings from the scientific literature to wide and diverse populations of people hinges upon the representativeness of the study sample with respect to demographic categories. The question to what extent the composition of a given study sample can make the generalisability of findings difficult or impossible has received attention in the field of psychology, where many articles published in prominent journals reflected participants from WEIRD (Western, Educated, Industrialized, Rich, and Democratic) contexts [@henrich_weirdest_2010; @muthukrishna_beyond_2020]. In other fields, analyses similar to the one in the present review have been published [@obryant_reporting_2004; @sifers_reporting_2002; @jones_reporting_2020], but to our knowledge, this review represents a first look at the inclusion, reporting and analysis of participant demographics in chronobiology and sleep research.

@@

There is convergent knowledge that health and sleep and circadian physiology are subject to large individual differences. Demographic variables provide on lens through which to understand individual differences, and importantly can provide insights into systemic disadvantages and inequities. In the clinical domain, the need to time therapy based on a patient's individual circadian rhythm has more recently become the focus of the emerging field of chronotherapy or chronotherapeutics [@adam_core_2019; @greco_personalized_2020; @dijk_novel_2020; @hill_optimizing_2020]. Understanding interindividual variability needs to become a key research area to understand the circadian and sleep physiology in the face of human diversity.

@@

We turn to possible limitations of this review and the included analyses and discuss how they might introduce bias in our findings. First, we consider the possibility that the article selection procedure may have introduced biases. Our review only concerned articles from a subset of eight specialized journals. As a consequence, the included articles were necessarily published in these journals, ignoring relevant articles published in other specialized journals (such as those included in the list of candidate journals), and articles published in other, including interdisciplinary journals. This raises the question to what extent we may have missed a section of the literature that would have been relevant to include here. As an alternative strategy, we considered randomly sampling a subset of chronobiology and sleep research articles produced by a general search (e.g. on search from "sleep OR chronobiology" on MEDLINE), but considered this to be too permissive. Our strategy of selecting a subset of candidate journals provided a reasonable trade-off, as well as sampling from a range of field-specific outlets.

Due to the non-uniform distribution of publication years of the included articles (Fig. 1), variables derived from published papers and visualized and/or analyzed by year will have varying uncertainty, with reported percentages from publications of the earlier years being most uncertain. The fact that early years are represented with fewer articles, however, is not a function of our data set, but of the exponential growth of scientific output [@powell_introduction_2017; @bornmann_growth_2015; @parolo_attention_2015].

## Towards standardized reporting of demographic variables: From checklists to schemas?

There are guidelines and/or checklists for standardizing reporting of participant characteristics, such as CONSORT [@schulz_consort_2010] or STROBE [@von_elm_strengthening_2007] (an extensive database for health research reporting guidelines is provided by the Equator Network, https://www.equator-network.org/). Some biomedical journals [e.g. @robinson_transparent_2017] specifically state demographic reporting requirements in the author instructions. Similarly, some organizations may make recommendations of specific reporting items for specific types of study questions [@veitch_cie_2020].

These guidelines and/or checklists are largely focused on *what* is reported and not *how* it is reported. There is, *a priori*, however, no reason to not develop and use a standardized and machine-readable schema for reporting participant characteristics. The FAIR principles state that data should be findable, accessible, interoperable and reusable [@wilkinson_fair_2016], and one way of realizing these criteria is the use of data schemas which could prescribe categories of data and common naming schemes for reporting participant characteristics. Importantly, however, "what gets counted counts" [@dignazio_data_2020], and it will be imperative to understand to what extent such data schema may be exclusionary, e.g., by enforcing gender binaries [@hyde_future_2019], and whether any specific demographic variable is truly important (following the principle of data minimization). A further central point to consider is the extent to which disaggregation by demographic variables could be used to do harm in some way [@dignazio_data_2020].

@@

This review provided a first look at the inclusion, reporting and analysis of demographic variables in the chronobiology and sleep research literature, considering >1000 articles across eight journals. We addressed the need to consider individual differences, as well as the dependence of sleep and circadian rhythms and demographic variables. Additionally, we outline an opportunity to improve reporting of participant-level characteristics using formalized data schemas.
